\textls[-0]{\thetitle}

### 2023-03-15 metatheory-condensed.tex
Add QIIT ref, Add Coquand Universe ref.

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities (for each judgement) have the form
\goodTerm{\Gm}{t_1 \equiv t_2}{T},\goodType{\Gm}{T_1 \equiv T_2}{},\goodCtx{\Gm_1 \equiv \Gm_2}{} ..., etc. We consider our syntax to be quotiented by these definition equalities, and thus each syntax piece is actually one (equivalent) class of syntax. Doing such quotient gives us convertibility for free---when we have definitionally equal $A \equiv B$, thanks to the quotient, the derivation of $t : A$ is the same as the derivation of $t : B$.\YZ{Since you are talking about quotienting, should you mention that the underlying logical framework is QIIT?}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

We use Tarski-style universes \cite{hofmann1997syntax},
which, unlike Russell-style ones,
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation to avoid clutter.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities (for each judgement) have the form
\goodTerm{\Gm}{t_1 \equiv t_2}{T},\goodType{\Gm}{T_1 \equiv T_2}{},\goodCtx{\Gm_1 \equiv \Gm_2}{} ..., etc. Following \citet{altkap2016}, we consider our syntax to be quotiented by these equalities, and thus each syntax piece is actually one (equivalent) class of syntax. Doing such quotient gives us convertibility for free---when we have definitionally equal $A \equiv B$, thanks to the quotient, the derivation of $t : A$ is the same as the derivation of $t : B$.\YZ{Since you are talking about quotienting, should you mention that the underlying logical framework is QIIT?}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

Following \citet{kaposi2019gluing}, we use \emph{Universe à la Coquand} \cite{coquand2013presheaf}. Unlike Russell-style ones, Coquand universes are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better due to its closedness to Tarski-style ones~\cite{luo2012notes}. 
% and compatibility with our algebraic formulation of MLTT~\cite{coquand2018canonicity}. % I am not sure if I want this sentence or not
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation to avoid clutter.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}
